---
title: "Conversion of GeoMx data to Seurat object"
author: "Diana Vera Cruz"
date: "2025-03-18"
output: html_notebook
---


## Introduction

A NanoStringsSet object can be converted to a Seurat object for further analysis or visualization. 
This is based in: 
(https://bioconductor.org/packages/release/bioc/vignettes/GeomxTools/inst/doc/GeomxSet_coercions.html)
but using a different dataset. 

## Requirements

This script uses the output GeoMx object from the QC script.


```{r}
## Libraries
library(GeomxTools)
library(Seurat)
library(harmony) # devtools::install_github("immunogenomics/harmony")


## Read tidy object: QC and filtered, but missing normalization. 
geomx <- readRDS('../env/tidy_geomx_obj.RDS')
## Normalization: Q3
geomx <- normalize(geomx, norm_method="quant", desiredQuantile = .75, toElt = "q_norm")


sc <- as.Seurat(geomx, normData = 'q_norm')
## If the data has Coordinates, include it here. 
sc <- as.Seurat(geomx, normData = 'q_norm', coordinates = c("ROI_Coordinate_X", "ROI_Coordinate_Y"))

## Assign an initial indentity to the cells.
Idents(sc) = 'region' ## Either the name of the column in metadata or the vector with values pers segment/ROI.
```


## Feature plots

### VlnPlots

You can use the VlnPlots, with any of the features per segment, either from metadata or genes.
```{r}
VlnPlot(sc, features = 'nCount_GeoMx', pt.size = 0.1, group.by = 'region') 

VlnPlot(sc, features = 'area', pt.size = 0.1) 

VlnPlot(sc, features = 'nCount_GeoMx', pt.size = 0.1, group.by = 'class') 

VlnPlot(sc, features = 'LYZ', group.by = 'region')
```

## Dimension reduction and variance.

### PCA
```{r pca}
sc <- FindVariableFeatures(sc)
sc <- ScaleData(sc)
sc <- RunPCA(sc, assay = "GeoMx", verbose = FALSE)
```

```{r}
## Variance explained by each PC. Keep 90% of variance in first 50 PC.
x = (sc@reductions$pca@stdev)^2 / sum((sc@reductions$pca@stdev)^2) 
which( cumsum(x) > 0.9)[1] ## PC that first includes more than 90% of variance.
## Screeplot -> SD per PC and biplots. 
ElbowPlot(sc, ndims = 50) + geom_vline(xintercept = which( cumsum(x) > 0.9)[1], linetype = 'dashed', color = 'orange')
PCAPlot(sc, group.by = 'region', shape.by = 'class')
PCAPlot(sc, dims =c(3,4), group.by = 'region', shape.by = 'class')
```


```{r}
pcs = seq_len(28)
sc <- FindNeighbors(sc, reduction = "pca", dims = pcs)
sc <- FindClusters(sc, verbose = FALSE)
sc <- RunUMAP(sc, reduction = "pca", dims = pcs)

DimPlot(sc, reduction = "umap", label = TRUE, group.by = "region", shape.by = 'class')
DimPlot(sc, reduction = "umap", label = TRUE, group.by = "slide_name")
```

## Spatial Feature Plots

In this case, we will showcase the spatial feature plots for the one slide, disease3.
```{r}

SpatialFeaturePlot(sc[,sc$slide_name == "disease3"], features = "nCount_GeoMx", pt.size.factor = 8)

SpatialFeaturePlot(sc[,sc$slide_name == "disease3"], features = "NegProbe_pct", pt.size.factor = 8)

SpatialFeaturePlot(sc[,sc$slide_name == "disease3"], features = "LYZ", pt.size.factor = 8)
```
Nanostrings also developed a tool for this: 
https://github.com/Nanostring-Biostats/SpatialOmicsOverlay

## Batch correction

You can use other tools like `harmony` to correct data and visualize the results.

```{r}
## Run harmony.
sc <- RunHarmony(sc, group.by.vars = "slide_name", assay.use = "GeoMx", plot_convergence = TRUE)

## Run clustering and UMAP.
sc <- FindNeighbors(sc, reduction = "harmony", dims = 1:20)
sc <- FindClusters(sc, resolution = 0.5)
sc <- RunUMAP(sc, reduction = "harmony", dims = 1:20)

DimPlot(sc, reduction = "umap", group.by = "slide_name")
DimPlot(sc, reduction = "umap", group.by = "region", shape.by = 'class')
```



