---
title: "Spatial Deconvolution"
subtitle: 'SpatialDecon'
author: "Diana Vera Cruz"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
---

## Goal

Spatial Deconvolution infers cell abundance per cell type in each segment, based on a reference matrix. This method does not incorporate the spatial context for a given segment, rather infers cell types abundances based on expression, and if nuclei counts are available, it can be used to adjust the cell type abundances.

## Steps:

You can start with a "NanoStringGeoMxSet" or a "Seurat" object.

```{r setup, include=FALSE}
library(tidyverse)
library(SpatialDecon)
library(GeomxTools)
library(ComplexHeatmap)

geomx = readRDS('../env/tidy_geomx_obj.RDS')

## Initial normalization: Q3, Upperquartile normalization.
geomx <- normalize(geomx, 
                     fromElt = "exprs",
                     norm_method = "quant", 
                     desiredQuantile = .75,
                     toElt = "q_norm")

## Add Assumed cell type: Combination of region + segment. 
protocolData(geomx)$celltype = with(sData(geomx),  
     case_when(segment == 'Geometric Segment' ~ 'Glomeruli', 
               segment == 'PanCK-' ~ 'Distal\nTubules', 
               TRUE ~ 'Proximal\nTubules'))

```

### 1. Calculate Background (Optional)

This requires the presence of Negative Probes in the dataset.

**Input** \* norm: Normalized expression matrix (Linear scale) \* probepool: Probeset names \* negnames: Negative control probeset names

```{r}
## Calculate background for data-set: Based in Negative Control Probes. For each probeset, one value in negnames. 
##Check the negative control probes in the dataset, NegProbe prefix expected for most probesets. 
dplyr::select( fData(geomx), Module, TargetName) %>% filter(grepl('NegProbe', TargetName))

bg = derive_GeoMx_background(norm = geomx@assayData$q_norm,
                             probepool = fData(geomx)$Module,
                             negnames = c("NegProbe-WTX"))
```

### 2. Cell Profile matrix

#### Predefined tables: Human/Mouse Atlas from Nanostring

```{r}
## Obtain the Kidney Dataset (Closest set that resembles our data)
## Obtained from https://github.com/Nanostring-Biostats/CellProfileLibrary/tree/master
human_kidney <- download_profile_matrix(species = "Human",
                                       age_group = "Adult", 
                                       matrixname = "Kidney_HCA")


heatmap(sweep(human_kidney, 1, apply(human_kidney, 1, max), "/"),
        labRow = NA, margins = c(10, 5))

#Heatmap(human_kidney, show_column_dend = F, show_row_dend = F, use_raster = T)
```

#### Custom reference matrix.

You can also create a custom reference matrix derived from a set you annotated previously. The functions for this are included in [Bioconductor: SpatialDecon_vignette](https://bioconductor.org/packages/release/bioc/vignettes/SpatialDecon/inst/doc/SpatialDecon_vignette_NSCLC.html){.uri}

### Run deconvolution

```{r}
res = runspatialdecon(
  object = geomx, 
  norm_elt = 'q_norm', 
  raw_elt = 'exprs',
  X = human_kidney, ## Reference matrix
  #cell_counts = sData(geomx)$nuclei ## Add if you have cell nuclei
)

beta_mx = res$beta

```

```{r, eval = F}
res = spatialdecon(
  norm = geomx@assayData$q_norm, 
  bg = bg,
  X = human_kidney, ## Reference matrix
  raw = geomx@assayData$exprs,
  cell_counts = sData(geomx)$nuclei 
)

beta_mx = res$beta
## Cell

```

```{r}
TIL_barplot(mat = t(beta_mx), draw_legend = T)
```

```{r}

Heatmap(t(beta_mx[,colSums(beta_mx) > 0]), show_column_names = F, show_row_dend = F,
        col = c('white', 'red3'), name = 'Cell\nabundance\nscore',
        show_parent_dend_line = T, row_names_side = 'left',
        #show_heatmap_legend = F, 
        row_names_gp = gpar(fontsize = 8), column_title_gp = gpar(fontsize = 10),
        top_annotation = HeatmapAnnotation(df = dplyr::select(sData(geomx), class), 
                                           col = list(class = c(DKD = 'purple', normal = 'grey'))),
        column_split = sData(geomx)$celltype)

```
